---
title: "Small benchmark of cell type isolation with scGate and other methods"
author: "M. Andreatta, A. Berenstein and S. Carmona"
date: "29/09/2021"
output:
  rmdformats::readthedown:
    self-contained: true
    highlight: haddock
    thumbnails: false
    css: styles.css
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'scGate.bench.html'))})
---


```{r, message=F, warning=F,results=F}
library(renv)
renv::activate()
renv::restore()
#remotes::install_github("carmonalab/UCell", ref="v1.1")
#remotes::install_github("mojaveazure/seurat-disk",ref="163f1aade5bac38ed1e9e9c912283a7e74781610")
#remotes::install_github("carmonalab/scGate", ref='v1.0.0') #stable release
#install.packages("SCINA")


#install.packages("BiocManager")
#BiocManager::install("devtools")
#library(devtools)
#BiocManager::install(c('DelayedArray', 'DelayedMatrixStats',"monocle","org.Hs.eg.db","org.Mm.eg.db"))
#devtools::install_github("cole-trapnell-lab/garnett")


library(ggplot2)
library(dplyr)
library(scGate)
library(SingleR)
library(SCINA)
library(garnett)
library(org.Hs.eg.db)

```


## Load some testing datasets and models

Load datasets (you will need ~ 14gb of RAM)
```{r}
rds.file = "./aux/human_sets_benchmark.rds"
if (!file.exists(rds.file)) {
  options(timeout=2000)
  download.file("https://figshare.com/ndownloader/files/31346815?private_link=7d4d9376a3bbdb17bc00", rds.file)
}
testing.datasets <- readRDS(rds.file) 

#ground_truth_categories <- readRDS("./aux/SingleR_goldStandard.rds")  #Load an object containing stardard cell type annotations - to allow comparing different methods
ground_truth_categories <- readRDS("./aux/goldStandards.rds")  #Load an object containing stardard cell type annotations - to allow comparing different methods

figure.path <- "./plots/"
dir.create(figure.path)

```

# Load pbmc datasets from Abdelaal 2019
(see processing steps in scGate_dev repo)
```{r}
runing_abdelaal = F
if(runing_abdelaal){
  testing.datasets <- readRDS(paste0("aux/pbmcs_abdelaal2019.rds"))
  platforms = c("10Xv2","10Xv3","CEL-Seq","Drop-Seq","inDrop","Seq-Well","Smart-Seq2")
}
```


Load default models from DB
```{r}
models.DB <- scGate::get_scGateDB(version="v0.4")  ### NB: force update will make results non-reproducible. We must specify a tag for download, or alternatively include the set of models in the repo

models <- models.DB$human$generic  ## Pre-compiled list of models for human cell types
print(models%>%names)

```


## Run models and compute performance

```{r}
## SingleR
library(celldex)
hpca.se <- HumanPrimaryCellAtlasData() 
param <- BiocParallel::MulticoreParam(workers = 4)

dset.to.run <- names(testing.datasets)
res.SingleR <- list()
SingleR.output <- list()
for(dset in dset.to.run){
    obj <- testing.datasets[[dset]]
    SingleR.output[[dset]] <- SingleR(test = GetAssayData(obj), ref = hpca.se,labels = hpca.se$label.fine, BPPARAM=param)
    res.SingleR[[dset]][["SingleR.fine"]] <- SingleR.output[[dset]]$pruned.labels
}
```


Run scGate
```{r,collapse =T, results= F, echo = F}
# List available testing datasets:
#names(testing.datasets)

results <- data.frame()
models <- models[names(models)!="Tcell.alphabeta"] #skip this model since there is no equivalent class in hpca for SingleR

for(dset in dset.to.run){
  message(sprintf('-Running %s',dset))
  obj <- testing.datasets[[dset]]
  assay <-  DefaultAssay(obj)
  

  for(model.name in names(models)){
    message(sprintf('---Running %s',model.name))
    model <- models[[model.name]]
    obj <- scGate(data = obj, model = model, ncores = 4, keep.ranks = TRUE, output.col.name = paste0('is.pure.',model.name), assay = assay)
    
    res <- scGate::performance.metrics(actual = obj@meta.data[,model.name], pred = obj@meta.data[,paste0("is.pure.",model.name)] == "Pure")
    res <- data.frame(value = res, metric = names(res),model = model.name,dataset = dset,method = "scGate")
    results <- rbind(results,res)
    
    #merge SingleR results
    pred.singleR <- (res.SingleR[[dset]])$SingleR.fine%>%grepl(ground_truth_categories$hpca_categories[[model.name]],.)+0
    singleR <- scGate::performance.metrics(actual = obj@meta.data[,model.name], pred = pred.singleR == 1)
    singleR <- data.frame(value = singleR, metric = names(singleR),model = model.name,dataset = dset,method = 'SingleR')
    results <- rbind(results,singleR)
  }

}
#saveRDS(results,"./aux/results.rds")
```

# Build SCINA model signatures 
```{r}
scina.signatures <- list()
scina.signatures$Bcell <- c("PTPRC","LCK","MS4A1","BANK1","PAX5","CD19","CD79A") 
scina.signatures$Plasma <- c("PTPRC","LCK","IGKC","IGHG3","IGHG1","IGHA1","CD79A") 
scina.signatures$PanBcell <- c("PTPRC","LCK","CD79A")
scina.signatures$MoMacDC <- c("PTPRC","SPI1","LYZ","CSF1R","MSR1","MAFB","CD300E")
scina.signatures$Neutrophils <- c("PTPRC","SPI1","CSF3R","FCGR3B")
scina.signatures$Myeloid <- c("PTPRC","SPI1")
scina.signatures$Tcell <- c("CD3D","CD3E","CD3G","CD2")
scina.signatures$CD4T <- c("CD3D","CD3E","CD3G","CD2","CD4","CD40LG")
scina.signatures$CD8T <- c("CD3D","CD3E","CD3G","CD2","CD8A","CD8B")
scina.signatures$Tgammadelta <- c("CD3D","CD3E","CD3G","CD2","TRDC","TRGC1","TRGC2","TRDV1")
scina.signatures$NK <- c("LCK","KLRD1","NKG7","NCR1","FCGR3A")
scina.signatures$Epithelial <- c("CDH1","FLT1")
scina.signatures$Stromal <- c("MMP2","COL1A1","COL1A2","COL5A1","LUM","PDGFRA")
scina.signatures$Erythrocyte <- c("HBB","HBA2","HBA1")
```

We cannot simply run SCINA with multiple, conflicting signatures, as the method would be forced to assign to one class or another (e.g. myeloid or MoMacDC). Instead, we need to set up a "model" for each target cell type that avoids conflicts.

```{r}
scina.model <- list()
scina.model$Bcell <- scina.signatures[c("Bcell","MoMacDC","Tcell","NK","Epithelial","Stromal","Erythrocyte")]
scina.model$PanBcell <- scina.signatures[c("PanBcell","MoMacDC","Tcell","NK","Epithelial","Stromal","Erythrocyte")]
scina.model$Plasma_cell <- scina.signatures[c("Plasma","MoMacDC","Tcell","NK","Epithelial","Stromal","Erythrocyte")]
scina.model$MoMacDC <- scina.signatures[c("MoMacDC","Bcell","Tcell","NK","Epithelial","Stromal","Erythrocyte")]
scina.model$Myeloid <- scina.signatures[c("Myeloid","Bcell","Tcell","NK","Epithelial","Stromal","Erythrocyte")]
scina.model$Tcell <- scina.signatures[c("Tcell","Bcell","MoMacDC","NK","Epithelial","Stromal","Erythrocyte")]
scina.model$CD4T <- scina.signatures[c("CD4T","Bcell","MoMacDC","CD8T","Tgammadelta","NK","Epithelial","Stromal","Erythrocyte")]
scina.model$CD8T <- scina.signatures[c("CD8T","Bcell","MoMacDC","CD4T","Tgammadelta","NK","Epithelial","Stromal","Erythrocyte")]
scina.model$NK <- scina.signatures[c("NK","Bcell","MoMacDC","Tcell","Epithelial","Stromal","Erythrocyte")]
```

# Run SCINA
```{r}
for(dset in dset.to.run){
  message(sprintf('-Running %s',dset))
  obj <- testing.datasets[[dset]]
  assay <-  DefaultAssay(obj)
  expr.mat <- GetAssayData(object = obj,assay = assay)
#  res.scina <- SCINA(expr.mat,signatures = scina.signatures, max_iter=100,convergence_n=10,convergence_rate=0.99,sensitivity_cutoff=0.9,rm_overlap=F)

  for(model.name in names(models)){
    
    res.scina <- SCINA(expr.mat,signatures = scina.model[[model.name]], rm_overlap = F)

    pred.SCINA <- (res.scina$cell_labels==model.name) + 0
    res.SCINA <- scGate::performance.metrics(actual = obj@meta.data[,model.name], pred = pred.SCINA == 1)
    res.SCINA <- data.frame(value = res.SCINA, metric = names(res.SCINA),model = model.name,dataset = dset,method = 'SCINA')
    results <- rbind(results,res.SCINA)
  }  
}
```



## Run Garnett pre-trained model (pbmc)
Download and load pre-trained classifier
```{r}
download.file("https://cole-trapnell-lab.github.io/garnett/classifiers/hsPBMC_20191017.RDS",destfile = "./aux/garnett_hsPBMC_classifier.RDS")
download.file("https://cole-trapnell-lab.github.io/garnett/marker_files/hsPBMC_markers.txt",destfile = "./aux/garnett_hsPBMC_markerList.txt")
pbmc_classifier <- readRDS("./aux/garnett_hsPBMC_classifier.RDS")  # load pre-trained classifier

```



# Garnet classification
```{r}
for(dset in dset.to.run){
  message(sprintf('-Running %s',dset))
  obj <- testing.datasets[[dset]]
  
  #adapt format
  cds.obj <- as.CellDataSet(obj)
  cds.obj <- estimateSizeFactors(cds.obj)

  #make classif
  cds.obj <- classify_cells(cds.obj, pbmc_classifier,
                             db = org.Hs.eg.db,
                             cluster_extend = TRUE,
                             cds_gene_id_type = "SYMBOL")
  

  for(model.name in names(models)){
    pred.garnett.cellType <- (pData(cds.obj)$cell_type) %>%grepl(ground_truth_categories$garnett.PBMC.clasif.pre.trained[[model.name]],.)+0
    res.garnett <- scGate::performance.metrics(actual = obj@meta.data[,model.name], pred = pred.garnett.cellType == 1)
    res.garnett <- data.frame(value = res.garnett, metric = names(res.garnett),model = model.name,dataset = dset,method = 'Garnett')
    results <- rbind(results,res.garnett)

    pred.garnett.ext.cellType <- (pData(cds.obj)$cluster_ext_type) %>%grepl(ground_truth_categories$garnett.PBMC.clasif.pre.trained[[model.name]],.)+0
    res.garnett.ext <- scGate::performance.metrics(actual = obj@meta.data[,model.name], pred = pred.garnett.ext.cellType == 1)
    res.garnett.ext <- data.frame(value = res.garnett.ext, metric = names(res.garnett.ext),model = model.name,dataset = dset,method = 'Garnett.clust.ext')
    results <- rbind(results,res.garnett.ext)

  }  
}
```


```{r}
saveRDS(results,"aux/benchmark.results.rds")
```





Summarize performance
```{r,fig.width = 8,fig.height=5}
data_summary <- function(x) {
     m <- mean(x)
     se <- function(x) sqrt(var(x)/length(x))
     ymin <- m-se(x)
     ymax <- m+se(x)
     return(c(y=m,ymin=ymin,ymax=ymax))
}
  
res <- results%>%subset(metric %in% c("PREC","MCC") & model!= "Plasma_cell")
  
  
res.prec <- res%>%subset(metric == "PREC")
res.mcc <- res%>%subset(metric == "MCC")
```


And visualize performance metrics in one plot
```{r}
colors <- c("#E69F00", "#56B4E9","#56E9BA")

res.prec <- res.prec[!is.na(res.prec$value),]
res.prec$model <- factor(res.prec$model, levels=c("Bcell","PanBcell","MoMacDC","Myeloid","NK","Tcell","CD4T","CD8T"))   #reorder in more intuitive way

smm.prec <- aggregate(res.prec$value, list(res.prec$method, res.prec$model), FUN=mean)
colnames(smm.prec) <- c("method","model","value")

plt.prec <- ggplot() +
  geom_bar(data=smm.prec, aes(x=model, y=value, fill=method), stat="identity", position="dodge", width = 0.75, alpha=0.3) +
  geom_jitter(data=res.prec, aes(x=model, y=value, color = method, shape = dataset), 
              position=position_jitterdodge(jitter.width = 0.1), size = 2) +
  scale_color_manual(values=colors) + scale_fill_manual(values=colors) +
  ylab("Precision")   + theme_bw() + theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.4, hjust=0.5)) 
  

res.mcc <- res.mcc[!is.na(res.mcc$value),]
res.mcc$model <- factor(res.mcc$model, levels=c("Bcell","PanBcell","MoMacDC","Myeloid","NK","Tcell","CD4T","CD8T"))   #reorder in more intuitive way

smm.mcc <- aggregate(res.mcc$value, list(res.mcc$method, res.mcc$model), FUN=mean)
colnames(smm.mcc) <- c("method","model","value")


plt.mcc <- ggplot() +
  geom_bar(data=smm.mcc, aes(x=model, y=value, fill=method), stat="identity", position="dodge", width = 0.75, alpha=0.3) +
  geom_jitter(data=res.mcc, aes(x=model, y=value, color = method, shape = dataset), 
              position=position_jitterdodge(jitter.width = 0.1), size = 2) +
  scale_color_manual(values=colors) + scale_fill_manual(values=colors) +
  ylab("MCC")   + theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.4, hjust=0.5)) 

a <- plt.prec | plt.mcc
b <- plt.prec / plt.mcc

a

ggsave(file.path(figure.path,"benchmark_15012022_scGate_lands.pdf"), plot=a, width=9, height=4)

#remove x axis labels
plt.prec <- plt.prec + theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

b <- plt.prec / plt.mcc
b

ggsave(file.path(figure.path,"benchmark_15012022_scGate_portr.pdf"), plot=b, width=6, height=4)

```

```{r,echo =FALSE, results=F}
saveRDS(results,file = file.path(figure.path,"benchmark_15012022_scGate.rds")) 
```


